Towards a Verified Bzip2: A Functional Specification for Burrows-Wheeler

### 2019-05-02 draft_paper/draft_paper.tex
Stash

The Burrows-Wheeler transform (BWT) is an invertible transformation
that makes a string more amenable to compression by other methods
\cite{bw}. The forwards transform is straightforward to describe: for
a string of length \(n\), form the \(n \times n\) rotaion matrix of all
cyclic shifts of the original string, sort it lexicographically, and
take the last column of the matrix. In addition, we note that index
where the original string appears in the sorted rotation matrix. As
shown in \cref{fig:bw_ex}, applying the algorithm to the string
\texttt{abracadabra!} produces the new string \texttt{ard!rcaaabb} and
the index \texttt{3}. The transformation tends to produce runs like
the three \texttt{a}s or the two \texttt{b}s, and it is this feature
that makes the resulting string easier to compress. Although it is not
at all obvious, the transformed string, together with the index,
provide enough information to recover the the original string. BWT is
discussed more in \cref{sec:bwt}.

\begin{figure}
  \centering
  \begin{tt}
  \begin{tabular}{rc}
    0  & abracadabra! \\
    1  & bracadabra!a \\
    2  & racadabra!ab \\
    3  & acadabra!abr \\
    4  & cadabra!abra \\
    5  & adabra!abrac \\
    6  & dabra!abraca \\
    7  & abra!abracad \\
    8  & bra!abracada \\
    9  & ra!abracadab \\
    10 & a!abracadabr \\
    11 & !abracadabra
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{rc}
    0  & !abracadabr\textbf{a} \\
    1  & a!abracadab\textbf{r} \\
    2  & abra!abraca\textbf{d} \\
    \textit{3} & \textit{abracadabra\textbf{!}} \\
    4  & acadabra!ab\textbf{r} \\
    5  & adabra!abra\textbf{c} \\
    6  & bra!abracad\textbf{a} \\
    7  & bracadabra!\textbf{a} \\
    8  & cadabra!abr\textbf{a} \\
    9  & dabra!abrac\textbf{a} \\
    10 & ra!abracada\textbf{b} \\
    11 & racadabra!a\textbf{b}
  \end{tabular}
  \end{tt}
  \caption{BWT acting on \texttt{abracadabra!}}
  \label{fig:bw_ex}
\end{figure}

@@

Our main contribution is a formalization in the Coq proof assistant of
the BWT, together with a machine-checked proof that our implementation
is correct. Specifically, we implement the functions
$\mathtt{bwp} : \mathtt{list}~A \to \mathtt{list}~A$ and
$\mathtt{bwn} : \mathtt{list}~A \to \mathbb{N}$ to produce the string
and index part of the BWT respectively, and the function
$\mathtt{unbwt} : \mathbb{N} \to \mathtt{list}~A \to \mathtt{list}~A$ to
reverse the transform, and we prove the following theorem:
\begin{theorem}[unbwt\_correct]
  \label{thm:unbwt_correct}
  For all lists $l$,
  \begin{equation*}
    \mathtt{unbwt}~(\mathtt{bwn}~l)~(\mathtt{bwp}~l) = l
  \end{equation*}
\end{theorem}
The implementation of the functions and the outline of the proof are
based on Bird et~al.'s derivations.

@@

As a part of the proof of \Cref{thm:unbwt_correct}, we implemented
radix-sort and proved its correctness,
\begin{theorem}[radixsort\_correct]
  \label{thm:radixsort_correct}%
  For all lists $l$,
  \begin{equation*}
    \mathtt{Sorted}~(\mathtt{radixsort}~l) \land \mathtt{Permutation}~(\mathtt{radixsort}~l)~l
  \end{equation*}
\end{theorem}
and proved that any two stable sorts are interchangable,
\begin{theorem}[stable\_sort\_unique]
  \label{thm:radixsort_correct}%
  For all functions $f, g : \mathtt{list}~A \to \mathtt{list}~A$,
  \begin{gather*}
    (\forall l,~\mathtt{Sorted}~(f~l) \land \mathtt{StablePermutation}~(f~l)~l) \implies \\
    (\forall l,~\mathtt{Sorted}~(g~l) \land \mathtt{StablePermutation}~(g~l)~l) \implies \\
    \forall l,~f~l = g~l
  \end{gather*}
\end{theorem}
This in turn required developing suitable definitions for
\texttt{Sorted}, \texttt{Permutation}, and \texttt{StablePermutation}.
We explored multiple definitions for each of these properties, and
proved equivalence between them.

The rest of this paper is organize as follows. \Cref{sec:bwt} details
the history of the BWT and its applications. \Cref{sec:ord} defines
what we need to know about a type $A$ to sort a $\mathtt{list}~A$ as a
part of BWT. Following Bird et~al., \cref{sec:forwards_BWT} defines
the implementation of the forwards BWT in Coq, and
\cref{sec:inverse_BWT} derives an implementation of the inverse BWT
via program calculation. \Cref{sec:sorted_perm_stable} provides
multiple definitions of \texttt{Sorted}, \texttt{Permutation}, and
\texttt{StablePermutation}, and proves equivalence between them.
\Cref{sec:sort} implements insertion sort and radixsort, and proves
their correctness. \Cref{sec:BWT_proof} more formally proves the
theorems about the BWT mentioned in previous sections, including
\Cref{thm:unbwt_correct}.

@@

We represent a total, decidable preorder on type with the
\texttt{Preord} typeclass in Coq:
\lstinputlisting[firstline=9, lastline=14]{../theories/BWT/Sorting/Ord.v}
We do not explicitly require reflexivity because it follows from totality.

@@

We can also define $x \equiv y \coloneqq x \le y \land y \le x$. $\equiv$ is a
decidable equivalence relation, which means it is reflexive,
transitive, and symmetric. The builtin class \texttt{EqDec} is used to
describe types with decidable equivalence relations, so for any
\texttt{Preord} instance we have the \texttt{EqDec} instance
\texttt{Preord\_EqDec} using \texttt{eqv\_dec}:
\lstinputlisting[firstline=116, lastline=120,
  widthgobble=2]{../theories/BWT/Sorting/Ord.v}
In Coq, the notation \texttt{===} is used for the relation of an
\texttt{EqDec} instance and \texttt{==} for the decision procedure of
an \texttt{EqDec} instance.

We say that we have a total, decidable \textit{order} when we have
that for all $x$ and $y$,
\begin{equation*}
  x \equiv y \implies x = y
\end{equation*}
that is, the induced equivalence relation of the preorder ($\equiv$)
implies Leibniz equality ($=$). We use the \texttt{Ord} class in Coq for this:
\lstinputlisting[firstline=175, lastline=176,
  widthgobble=2]{../theories/BWT/Sorting/Ord.v}
For the BWT, we require an \texttt{Ord} instance for $A$, but for other parts of the proof, such as sorting, \texttt{Preord} will suffice.

@@

First, we define left and right rotations of a list.
\lstinputlisting[widthgobble=2, firstline=17, lastline=22]{../theories/BWT/Lib/List.v}
\lstinputlisting[widthgobble=2, firstline=16, lastline=21]{../theories/BWT/Rotation/Rotation.v}
\lstinputlisting[widthgobble=2, firstline=23, lastline=28]{../theories/BWT/Rotation/Rotation.v}

Now we can produce the rotation matrix by repeatedly applying \texttt{lrot}.
\lstinputlisting[widthgobble=2, firstline=11, lastline=18]{../theories/BWT/Lib/Iterate.v}
\lstinputlisting[widthgobble=2, firstline=17, lastline=18]{../theories/BWT/Rotation/Rots.v}

Finally, we produce the transformed string by taking the last column
of the sorted rotation matrix. \texttt{lexsort} sorts the rotation
matrix lexicographically using the \texttt{Ord} instance; we will
discuss it's implementation in \cref{sec:sort}.
\lstinputlisting[firstline=31,
  lastline=35]{../theories/BWT/BurrowsWheeler.v}
Note that in order to ensure that $\mathtt{last}$ is total, it takes
an extra argument to return in the case of an empty list.

We also need to find the index in the sorted rotation matrix where the
original string appears.
\lstinputlisting[firstline=12, lastline=19]{../theories/BWT/Lib/FindIndex.v}
\lstinputlisting[firstline=73, lastline=74]{../theories/BWT/BurrowsWheeler.v}

@@

Now we have to construct a function \texttt{unbwt} such that for all $l$,
\begin{equation}
  \mathtt{unbwt}~(\mathtt{bwn}~l)~(\mathtt{bwp}~xs) = l
  \label{eq:unbwt}
\end{equation}
If we could recreate the entire sorted rotation matrix from
$\mathtt{bwp}~l$, it would be simple to recover $x$ by indexing into
the matrix at $\mathtt{bwn}~x$.

More precisely, suppose we had a function function \texttt{recreate}
such that
\begin{equation}
  \mathtt{recreate}~(\mathtt{bwp}~l) = \mathtt{lexsort}~(\mathtt{rots}~l)
  \label{eq:recreate}
\end{equation}
Then, we could define \texttt{unbwt} as
\begin{lstlisting}
Definition unbwt (i : nat) (l : list A) : list A :=
  nth i (recreate l) l.
\end{lstlisting}
and \Cref{eq:unbwt} would follow directly from the following theorem:
\begin{theorem}[bwn\_correct]
  For every non-empty list $l$,
  \begin{equation*}
    (\mathtt{lexsort}~(\mathtt{rots}~l))_{(\mathtt{bwn}~l)} = l
  \end{equation*}
\end{theorem}

How, then, do we implement \texttt{recreate}? We will recreate the
sorted rotation matrix column by column, so \verb|recreate j| will
recreate the first \verb|j| columns of the sorted rotation matrix.
\begin{verbatim}
Definition cols (j : nat) (l : list A) : list A := map (firstn j) l.

Theorem recreate_correct_ind : forall j l,
    j <= length l ->
    recreate j (bwp l) = cols j (lexsort (rots l)).
\end{verbatim}

To derive an implementation of \verb|recreate| from this correctness
specification, we need an expression for \verb|recreate 0| and
\verb|recreate (S j)|. We can substitute for \verb|j| on the
right-hand side and manipulate that expression algebraically until it
is in the form \verb|_ (bwp l)|. Then we can define \verb|recreate 0|
and \verb|recreate (S j)| to be precisely the \verb|_|, and by
running our derivation in reverse, we obtain a proof for the base and
inductive cases that uses only rewrites.

We start with the base case, \verb|recreate 0|:
\begin{verbatim}
  cols 0 (lexsort (rots l))
= { by unfold cols }
  map (firstn 0) (lexsort (rots l))
= { by firstn_O}
  map (const []) (lexsort (rots l))
= { by map_const }
  repeat [] (length (lexsort (rots l)))
= { by bwp_length, rots_length, lexsort_length }
  repeat [] (length (bwp l))
= { by map_const }
  map (const []) (bwp l)
\end{verbatim}

From this, we can define \verb|recreate 0 l = map (const []) l| and
we already have the proof for the base case of the correctness
specification.

For the inductive case, we have the expression
\begin{verbatim}
  cols (S j) (lexsort (rots l))
\end{verbatim}
on the right-hand side. Our plan is to express \verb|cols (S j)|
in terms of \verb|cols j|, and then to use the inductive hypothesis
\begin{verbatim}
  cols j (lexsort (rots l)) = recreate j
\end{verbatim}
to produce a well-formed recursive definition.

One way to do this is to say that taking the first \verb|S j| columns
of a matrix is the same as moving the first column to the back, taking
the first \verb|j| columns, and then sticking the last column back on
the front:
\begin{verbatim}
  cols (S j) l = prepend_col (map last (map lrot l)) (cols j (map lrot l))
\end{verbatim}
If we substitute \verb|map rrot l| for \verb|l| we find a simpler
expression
\begin{verbatim}
  cols (S j) (map rrot l) = prepend_col (map last l) (cols j l)
\end{verbatim}

Now we would like to find a way of introducing a \verb|map rrot|
before the \verb|lexsort (rots l)|. The insight here is that we can
assume that \verb|lexsort| is actually a radix-sort. That is, that
\begin{verbatim}
lexsort xs = rep (length xs) (hdsort ∘ map rrot) xs
\end{verbatim}
where \verb|hdsort| sorts the matrix stably on the the first column
and \verb|rep| is defined as
\begin{verbatim}
Fixpoint rep (f : A -> A) (n : nat) (z : A) : A :=
  match n with
  | O => z
  | S m => f (rep f m z)
  end.
\end{verbatim}

We can do this because there is only one way to sort the matrix, so
assuming \verb|lexsort| and the radix-sort are correct, they must be
doing the same thing. We will specify precisely when and why this is
valid later.

Stepping back, it is important to say that this insight, whether
explicitly stated or not, is always the crucial step in inverting the
Burrows-Wheeler transform. Because radix-sort only needs one column at
a time, this transformation will allow us to reconstruct the sorted
rotation matrix one column at a time.

In seeking to introduce an extra \verb|map rrot| term, we can note
that rotating and sorting an extra time doesn't change anything when
we are sorting the rotation matrix. This works because rotating the
last column to the front produces a matrix that is sorted on its last
\verb|n - 1| characters, so sorting stably on the first column will
re-sort the matrix. Crucially though, rotating the last column
forwards also permutes the entire matrix, so in addition to having a
sorted list, we have a sorted list of the same elements.

More precisely, we can characterize how \verb|map rrot| permutes
\verb|rots|
\begin{verbatim}
map rrot (rots x) = rrot (rots x)
\end{verbatim}

Using that we can show
\begin{verbatim}
  lexsort (rots x)
= { because rrot permutes its elements }
  lexsort (rrot (rots x))
= { because hdsort permutes its elements }
  lexsort (hdsort (rrot (rots x)))
= { radix-sort }
  rep n (hdsort ∘ rrot) (hdsort (rrot (rots x)))
= { rep_absorb_r }
  rep (S n) (hdsort ∘ rrot) (rots x)
= { rep_split_l }
  hdsort (rrot (rep n (hdsort ∘ rrot) (rots x)))
= { radix-sort }
  hdsort (rrot (lexsort (rots x)))
\end{verbatim}

From here we can calculate
\begin{verbatim}
  cols (S j) (lexsort (rots l))
= cols (S j) (hdsort (map rrot (lexsort (rots l))))
= { hdsort commutes with cols (S j) }
  hdsort (cols (S j) (map rrot (lexsort (rots l))))
= { reduce cols (S j) }
  hdsort (prepend_col (map last (lexsort (rots l))) (cols j (lexsort (rots l)))
= { by IH and def of bwp }
  hdsort (prepend_col (bwp l) (recreate j (bwp l)))
= (fun x => hdsort (prepend_col x (recreate j x))) bwp
\end{verbatim}

We can use this to define \verb|recreate|:
\begin{verbatim}
Fixpoint recreate (j : nat) (l : list A) : list (list A) :=
  match j with
  | O    => map (const []) l
  | S j' => hdsort (prepend_col l (recreate j' l))
  end.
\end{verbatim}

\section{\texttt{Sorted}, \texttt{Permutation}, and \texttt{StablePermutation}}
\label{sec:sorted_perm_stable}

In this section we prove the lemmas from the previous section:
\begin{enumerate}
\item
\begin{verbatim}
map rrot (rots x) = rrot (rots x)
\end{verbatim}

\item \verb|hdsort| commutes with \verb|cols (S j)|
  This proof is cool because in theory it can be proven via
  parametricity, but that's a metatheorem in Coq.

\item
\begin{verbatim}
cols (S j) (map rrot l) = prepend_col (map last l) (cols j l)
\end{verbatim}
\end{enumerate}

>>>>

The Burrows-Wheeler transform (BWT) is an invertible transformation
that makes a string more amenable to compression by other methods
\cite{bw}. The forwards transform is straightforward to describe: for
a string of length \(n\), form the \(n \times n\) rotaion matrix of all
cyclic shifts of the original string, sort it lexicographically, and
take the last column of the matrix. In addition, we note the index
where the original string appears in the sorted rotation matrix. As
shown in \Cref{fig:bw_ex}, applying the algorithm to the string
\var{abracadabra!} produces the new string \var{ard!rcaaabb} and
the index \var{3}. The transformation tends to produce runs like
the three \var{a}s or the two \var{b}s in \var{ard!rcaaabb},
and it is this feature that makes the resulting string easier to
compress. Although it is not obvious, the transformed string, together
with the index, provide enough information to recover the the original
string. The BWT is discussed more in \cref{sec:bwt}.

\begin{figure}
  \centering
  \begin{tt}
  \begin{tabular}{rc}
    0  & abracadabra! \\
    1  & bracadabra!a \\
    2  & racadabra!ab \\
    3  & acadabra!abr \\
    4  & cadabra!abra \\
    5  & adabra!abrac \\
    6  & dabra!abraca \\
    7  & abra!abracad \\
    8  & bra!abracada \\
    9  & ra!abracadab \\
    10 & a!abracadabr \\
    11 & !abracadabra
  \end{tabular}
  $\rightarrow$
  \begin{tabular}{rc}
    0  & !abracadabr\textbf{a} \\
    1  & a!abracadab\textbf{r} \\
    2  & abra!abraca\textbf{d} \\
    \textit{3} & \textit{abracadabra\textbf{!}} \\
    4  & acadabra!ab\textbf{r} \\
    5  & adabra!abra\textbf{c} \\
    6  & bra!abracad\textbf{a} \\
    7  & bracadabra!\textbf{a} \\
    8  & cadabra!abr\textbf{a} \\
    9  & dabra!abrac\textbf{a} \\
    10 & ra!abracada\textbf{b} \\
    11 & racadabra!a\textbf{b}
  \end{tabular}
  \end{tt}
  \caption{BWT acting on \var{abracadabra!}}
  \label{fig:bw_ex}
\end{figure}

@@

Our main contribution is a formalization in the Coq proof assistant of
the BWT, together with a machine-checked proof that our implementation
is correct. Specifically, we implement the functions $\var{bwp} :
\var{list}~A \to \var{list}~A$ and $\var{bwn} : \var{list}~A
\to \mathbb{N}$ that produce the string and index part of the BWT
respectively, and the function $\var{unbwt} : \mathbb{N} \to
\var{list}~A \to \var{list}~A$ that reverses the transform. We
prove the following theorem:
\begin{theorem}[unbwt\_correct]
  \label{thm:unbwt_correct}
  For all lists $l$,
  \begin{equation*}
    \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
  \end{equation*}
\end{theorem}
The implementation of the functions and the outline of the proof are
based on Bird et~al.'s derivations.

@@

As a part of the proof of \Cref{thm:unbwt_correct}, we implemented
radix-sort and proved its correctness,
\begin{theorem}[radixsort\_correct]
  \label{thm:radixsort_correct}%
  For all lists $l$,
  \begin{equation*}
    \var{Sorted}~(\var{radixsort}~l) \land \var{Permutation}~(\var{radixsort}~l)~l
  \end{equation*}
\end{theorem}
and proved that any two stable sorts are interchangable,
\begin{theorem}[stable\_sort\_unique]
  \label{thm:radixsort_correct}%
  For all functions $f, g : \var{list}~A \to \var{list}~A$,
  \begin{gather*}
    (\forall l,~\var{Sorted}~(f~l) \land \var{StablePermutation}~(f~l)~l) \implies \\
    (\forall l,~\var{Sorted}~(g~l) \land \var{StablePermutation}~(g~l)~l) \implies \\
    \forall l,~f~l = g~l
  \end{gather*}
\end{theorem}
This in turn required developing suitable definitions for
\var{Sorted}, \var{Permutation}, and \var{StablePermutation}.
We explored multiple definitions for each of these properties, and
proved equivalence between them.

The rest of this paper is organize as follows. \Cref{sec:bwt} details
the history of the BWT and its applications. \Cref{sec:ord} defines
what we need to know about a type $A$ to sort a $\var{list}~A$ as a
part of BWT. Following Bird et~al., \cref{sec:forwards_BWT} defines
the implementation of the forwards BWT in Coq, and
\cref{sec:inverse_BWT} derives an implementation of the inverse BWT
via program calculation. \Cref{sec:sorted_perm_stable} provides
multiple definitions of \var{Sorted}, \var{Permutation}, and
\var{StablePermutation}, and proves equivalence between them.
\Cref{sec:sort} implements insertion sort and radixsort, and proves
their correctness. \Cref{sec:BWT_proof} more formally proves the
theorems about the BWT mentioned in previous sections, including
\Cref{thm:unbwt_correct}.

@@

We represent a total, decidable preorder on type with the
\var{Preord} typeclass in Coq:
\lstinputlisting[firstline=9, lastline=14]{../theories/BWT/Sorting/Ord.v}
We do not explicitly require reflexivity because it follows from
totality. Given a \var{Preord} instance for a type $A$, we will be
able to sort $\var{list}~A$ as shown in \cref{sec:sort}.

@@

We can also define $x \equiv y \coloneqq x \le y \land y \le x$. $\equiv$ is a
decidable equivalence relation, which means it is reflexive,
transitive, and symmetric, and we have a decision procedure for it.
\lstinputlisting[
  firstline=116, lastline=120, widthgobble=2
]{../theories/BWT/Sorting/Ord.v}

The builtin class \var{EqDec} is used to describe types with
decidable equivalence relations, so for any \var{Preord} instance
we have the \var{EqDec} instance \var{Preord\_EqDec} using
\var{eqv\_dec}. In Coq, the notation \var{===} is used for the
equivalence relation of an \var{EqDec} instance and \var{==} for
its decision procedure.

We say that we have a total, decidable \textit{order} when we have
that for all $x$ and $y$,
\begin{equation*}
  x \equiv y \implies x = y
\end{equation*}
that is, the induced equivalence relation of the preorder ($\equiv$)
implies Leibniz equality ($=$). We use the \var{Ord} class in Coq
for this:
\lstinputlisting[firstline=175, lastline=176, widthgobble=2]{../theories/BWT/Sorting/Ord.v}
For the BWT, we require an \var{Ord} instance for $A$, but
elsewhere, such as in sorting, \var{Preord} will suffice.

@@

First, we define the left and right rotations of a list. \var{lrot}
moves the first element of a list to the end.
\lstinputlisting[widthgobble=2, firstline=24, lastline=28]{../theories/BWT/Rotation/Rotation.v}
\var{rrot} moves the last element of a list to the front, and
relies on \var{init}, which drops the last element of a list.
\lstinputlisting[widthgobble=2, firstline=18, lastline=22]{../theories/BWT/Lib/List.v}
\lstinputlisting[widthgobble=2, firstline=17, lastline=21]{../theories/BWT/Rotation/Rotation.v}
Note that in order to ensure that $\var{last}$ is total, it takes
an extra argument to return in the case of an empty list.

Now we would likel to produce the rotation matrix by repeatedly
applying \var{lrot}. We will use \var{iter}, which takes a function
$f : A \to A$, a number $n$, and an initial value $z$, and returns the
list of intermediate values obtained by applying $f$ to $z$ $n$ times.
Or,
\begin{equation}
  \var{iter}~f~n~z = [z,~f~z,~f~(f~z),~\ldots,~f^n z]
  \label{eq:iter}
\end{equation}
\lstinputlisting[
  widthgobble=2,
  firstline=14,
  lastline=18
]{../theories/BWT/Lib/Iterate.v}

We can prove that this definition satisfies \Cref{eq:iter} by defining
the function \var{rep} for repeated function application.
\lstinputlisting[
  widthgobble=2,
  firstline=8,
  lastline=12
]{../theories/BWT/Lib/Repeat.v}
In proofs, we will still use the notation $f^i z$ for $\var{rep}~f~i~z$.

\begin{theorem}[rep\_r]
  For all $f$, $n$, and $z$,
  \begin{equation}
     f^i~(f~z) = f^{i+1}~z
  \end{equation}
\end{theorem}
\begin{proof}
  The theorem follows by induction on $n$.
\end{proof}

\begin{theorem}[iter\_length]
  For all $f$, $n$, and $z$,
  \begin{equation}
    \var{length}~(\var{iter}~f~n~z) = n
  \end{equation}
\end{theorem}
\begin{proof}
  The theorem follows by induction on $n$.
\end{proof}

\begin{theorem}[iter\_nth]
  For all $f$, $n$, $z$, and $i < n$,
  \begin{equation}
    (\var{iter}~f~n~z)_i = f^i~z
  \end{equation}
\end{theorem}
\begin{proof}
  The theorem follows by induction on $i$. The inductive case is
  \begin{align*}
       (\var{iter}~f~(n+1)~z)_{i+1}
    &= (z :: \var{iter}~f~n~(f~z))_{i+1} &&\{\text{\var{iter} def.}\} \\
    &= (\var{iter}~f~n~(f~z))_i          &&\{\text{by \var{nth} def.}\} \\
    &= f^i~(f~z)                            &&\{\text{by \var{rep\_r}}\} \\
    &= f^{i+1}~z && \qedhere
  \end{align*}
\end{proof}

Now we can define \var{rots} to produce the rotation matrix.
\lstinputlisting[
  widthgobble=2,
  firstline=17,
  lastline=18,
]{../theories/BWT/Rotation/Rots.v}

\var{bwp} produces the transformed string by taking the last column
of the sorted rotation matrix. \var{lexsort} sorts the rotation
matrix lexicographically using the \var{Ord} instance; we will
discuss it's implementation in \cref{sec:sort}.
\lstinputlisting[
  widthgobble=2,
  firstline=31,
  lastline=35
]{../theories/BWT/BurrowsWheeler.v}

\var{bwn} returns the index in the sorted rotation matrix where the
original string appears. It relies on \var{findIndex}, which is
defined with respect to some decidable equivalence relation.
\lstinputlisting[
  widthgobble=2,
  firstline=12,
  lastline=19,
]{../theories/BWT/Lib/FindIndex.v}

When we use \var{findIndex} in \var{bwn}, we are using the
induced equivalence relation from the \var{Ord} instance, so it
finds the row that is Leibniz equal to $l$.
\lstinputlisting[
  widthgobble=2,
  firstline=73,
  lastline=74,
]{../theories/BWT/BurrowsWheeler.v}

Given these definitions, we can quickly prove that \var{bwn}
satisfies its specification, namely that for all non-empty $l$
\begin{equation*}
    (\var{lexsort}~(\var{rots}~l))_{(\var{bwn}~l)} = l
\end{equation*}

\begin{lemma}[findIndex\_correct]
  For all elements $x$ and lists $l$,
  \begin{equation*}
    (\exists y \in l,~x \equiv y) \implies l_{(\var{findIndex}~x~xs)} \equiv x
  \end{equation*}
  \begin{proof}
    This follows by induction on $l$: At every step, either the
    current element is equivalent to $x$ and we return $0$, or there
    must be such a $y$ in the rest of the list.
  \end{proof}
\end{lemma}

\begin{lemma}[orig\_in\_sorted\_rots]
  For every non-empty list $l$,
  \begin{equation*}
    \exists y \in \var{lexsort}~(\var{rots}~l),~x = y
  \end{equation*}
  \begin{proof}
    Clearly $l$ is in $\var{rots}~l$ at index 0, and because
    \var{lexsort} produces a permutation of its input, $l$ must be
    in $\var{lexsort}~(\var{rots}~l)$ as well.
  \end{proof}
\end{lemma}

\begin{theorem}[bwn\_correct]
  For every non-empty list $xs$,
  \begin{equation*}
    (\var{lexsort}~(\var{rots}~xs))_{(\var{bwn}~xs)} = xs
  \end{equation*}
  \begin{proof}
    This follows from \var{findIndex\_correct} if we can show that
    $\exists y \in xs,~x = y$, which we have by \var{orig\_in\_sorted\_rots}.
  \end{proof}
\end{theorem}

@@

Now we want to construct a function \var{unbwt} such that for all $l$,
\begin{equation}
  \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
  \label{eq:unbwt}
\end{equation}
If we could recreate the entire sorted rotation matrix from
$\var{bwp}~l$, it would be simple to recover $l$ by indexing into
the matrix at $\var{bwn}~l$.

More precisely, suppose we had a function function \var{recreate}
such that
\begin{equation}
  \var{recreate}~(\var{bwp}~l) = \var{lexsort}~(\var{rots}~l)
  \label{eq:recreate}
\end{equation}
Then, we could define \var{unbwt} as
\begin{lstlisting}
Definition unbwt (i : nat) (l : list A) : list A :=
  nth i (recreate l) l.
\end{lstlisting}
and \Cref{eq:unbwt} would follow directly from the \var{bwn\_correct}:
\begin{align}
     \label{eq:unbwt_correct_start}
     \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l)
  &= (\var{recreate}~(\var{bwp}~l))_{(\var{bwn})~l} &&\\
  &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l}
     &&\text{\{by \Cref{eq:recreate}\}}\\
  &= l &&\text{\{\var{recreate\_correct}\}}
     \label{eq:unbwt_correct_end}
\end{align}

How, then, do we implement \var{recreate} such that
\Cref{eq:recreate} holds? One idea is to recreate the sorted rotation
matrix column by column. We define $\var{cols}~j$ to take the first
$j$ columns of a matrix:
\lstinputlisting[
  widthgobble=2,
  firstline=19,
  lastline=20,
]{../theories/BWT/Columns.v}
Then we introduce a parameter $j$ to \var{recreate}, and our new
specification is that $\var{recreate}~j$ recreates the first $j$
columns of the sorted rotation matrix:
\begin{theorem}[recreate\_correct\_ind]
  For all lists $l$ and $j \le \var{length}~l$,
  \begin{equation}
    \var{recreate}~j~(\var{bwp}~l) =
    \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
    \label{eq:recreate_correct_ind}
  \end{equation}
\end{theorem}

Now we can redefine \var{unbwt} as
\lstinputlisting[
  widthgobble=2,
  firstline=196,
  lastline=197,
]{../theories/BWT/BurrowsWheeler.v}

Assuming that we can prove \var{recreate\_correct\_ind}, the
correctness of of \text{unbwt} follows:

\begin{theorem}[cols\_id]
  For all $j$ and matrices $m : \var{list}~(\var{list}~A)$,
  \begin{gather*}
    (\forall r \in m,~\var{length}~r \le n) \implies \\
    \var{cols}~n~m = m
  \end{gather*}
\end{theorem}
\begin{proof}
  $\var{cols}~j~m = \var{map}~(\var{firstn}~j)~m$, and we
  know that $m = \var{map}~(\lambda x.x)~m$. So, it suffices to show that
  for every $r \in m$, the mapping functions agree, that is
  $\var{firstn}~j~r = r$. This follows from \var{firstn\_all2}
  in the Coq standard library.
\end{proof}

\begin{theorem}[iter\_preserves]
  We are given $f$, $n$, $z$ and a predicate $P : A \to \mathbb{P}$.
  Suppose $f$ preserves $P$, which is to say that
  \begin{equation*}
    \forall x,~P~x \implies P~(f~x)
  \end{equation*}
  then,
  \begin{gather*}
     P~z \implies \forall x \in \var{iter}~f~n~z,~P~x
  \end{gather*}
\end{theorem}
\begin{proof}
  This follows from induction on $n$.
\end{proof}

\begin{lemma}[sort\_rots\_all\_len]
  For all lists $l$,
  \begin{equation*}
    \forall x \in \var{lexsort}~(\var{rots}~l),~\var{length}~x = \var{length}~l
  \end{equation*}
\end{lemma}
\begin{proof}
  Because \var{lexsort} permutes its argument, it suffices to show
  that all the rows of $\var{rots}~l$ have the same length as $l$.
  This follows from \var{iter\_preserves} and the fact that
  \var{lrot} preserves length.
\end{proof}

\begin{corollary}[recreate\_correct]
  For all lists $l$,
  \begin{equation*}
    \var{recreate}~(\var{length}~l)~(\var{bwp}~l) =
    \var{lexsort}~(\var{rots}~l)
  \end{equation*}
\end{corollary}
\begin{proof}
  This follows from \var{recreate\_correct\_ind},
  \var{cols\_id}, and \var{sort\_rots\_all\_len}.
\end{proof}

\begin{theorem}[unbwt\_correct]
  For all lists $l$,
  \begin{equation*}
    \var{unbwt}~(\var{bwn}~l)~(\var{bwp}~l) = l
  \end{equation*}
\end{theorem}
\begin{proof}
  The proof is almost identical to that given in
  \Crefrange{eq:unbwt_correct_start}{eq:unbwt_correct_end}.
\end{proof}

To implement \var{recreate}, we need expressions for
$\var{recreate}~0$ and $\var{recreate}~(j+1)$ that satisfy
\Cref{eq:recreate_correct_ind} in \var{recrate\_correct\_ind}:
\begin{equation}
  \var{recreate}~j~(\var{bwp}~l) =
  \var{cols}~j~(\var{lexsort}~(\var{rots}~l))
  \tag{\ref{eq:recreate_correct_ind}}
\end{equation}

To derive an implementation, we can substitute for $j$ on the right-hand
side of \Cref{eq:recreate_correct_ind} and manipulate that expression
algebraically until it is in the form \var{\_ (bwp l)}. Then we can
define $\var{recreate}~0$ and $\var{recreate}~(j+1)$ to be
precisely those \var{\_}s, and by running our derivation in
reverse, we obtain a proof.

We start with the base case, $\var{recreate}~0$:
\begin{align*}
     \var{cols}~0~(\var{lexsort}~(\var{rots}~l))
  &= \var{map}~(\lambda x.[])~(\var{lexsort}~(\var{rots}~l))) \\
  &= \var{repeat}~[]~(\var{length}~(\var{lexsort}~(\var{rots}~l))) \\
  &= \var{repeat}~[]~(\var{length}~(\var{bwp}~l)) \\
     \var{cols}~0~(\var{lexsort}~(\var{rots}~l))
  &= \var{map}~(\lambda x.[])~(\var{bwp}~l)
\end{align*}

From this, we can define $\var{recreate}~0~l = \var{map}~(\lambda
x.[])~l$ and we already have the proof for the base case of the
correctness specification.

For the inductive case, we have the expression
\begin{equation*}
  \var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l))
\end{equation*}
on the right-hand side. Our plan is to express $\var{cols}~(j+1)$
in terms of $\var{cols}~j$, and then to use the induction
hypothesis
\begin{equation*}
  \var{cols}~j~(\var{lexsort}~(\var{rots}~l)) = \var{recreate}~j
\end{equation*}
to produce a well-formed recursive definition.

One way to do this is to note that taking the first $j+1$ columns of a
matrix is the same as moving the first column to the back, taking the
first $j$ columns, and then sticking the last column back on the
front:
\begin{equation*}
  \var{cols}~(j+1)~l =
  \var{prepend\_col}~(\var{map}~\var{hd}~l)~(\var{cols}~j~(\var{map}~\var{lrot}~l))
\end{equation*}
If we substitute $\var{map}~\var{rrot}~l$ for $l$ we find a simpler expression
\begin{equation}
  \label{eq:cols_rrot}
  \var{cols}~(j+1)~(\var{map}~\var{rrot}~l) =
  \var{prepend\_col}~(\var{map}~\var{last}~l)~(\var{cols}~j~l)
\end{equation}

Now we would like to find a way of introducing a $\var{map}~\var{rrot}$
before the $\var{lexsort}~(\var{rots}~l)$. The insight here is that we can
assume that \var{lexsort} is actually a radix-sort. That is, that
\begin{equation*}
\var{lexsort}~l = (\var{hdsort} \comp \var{map}~\var{rrot})^{\var{length}~l}~l
\end{equation*}
where \var{hdsort} sorts the matrix stably on the the first column.

We can do this because there is only one way to sort the matrix, so
assuming \var{lexsort} and the radix-sort are correct, they must be
doing the same thing. We will prove this in
\cref{sec:sorted_perm_stable,sec:sort}.

%% In seeking to introduce an extra \verb|map rrot| term, we can note
%% that rotating and sorting an extra time doesn't change anything when
%% we are sorting the rotation matrix. This works because rotating the
%% last column to the front produces a matrix that is sorted on its last
%% \verb|n - 1| characters, so sorting stably on the first column will
%% re-sort the matrix. Crucially though, rotating the last column
%% forwards also permutes the entire matrix, so in addition to having a
%% sorted list, we have a sorted list of the same elements.
